---
title: "Validation and survival analysis HLA-II^CONST signature"
author: "Mario Presti"
date: '`r paste("First created on November 2024 Updated on ", format(Sys.Date(), "%d %B %Y"))`'
output:
  github_document:
    toc: true
    toc_depth: 5
knit: |
  (function(inputFile, encoding) {
    dir.create("../Reports", showWarnings = FALSE)
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../Reports")
  })
---

```{r setup, include=FALSE}
library(biomaRt)
library(TCGAbiolinks)
library(singscore)
library(survival)
library(survminer)
library(DESeq2)
library(SummarizedExperiment)
library(edgeR)
library(dplyr)
library(openxlsx)
library(ggsurvfit)
library(ComplexHeatmap)
library(tidyr)
library(msigdbr)
library(clusterProfiler)
library(progeny)
library(GSVA)
library(binom)
library(ggpubr)
library(GSVA)
library(dplyr)
library(pheatmap)
library(stringr)
library(immunarch)

load("../controlled_data/combined_dds.RData")
```

# SIGNATURE CREATION

With the top/bottom 20 genes, a gene signature was created to investigate survival in TCGA patient.
Of note: baseMean above 3000 was chosen as a cutoff as we wanted to use highly expressed genes to investigate tumor biopsies, T cell and immune-related genes were excluded from the signature as in Gokuldass et al, Cancer Immunol Immunoth. 2022
#create signature
```{r, eval=TRUE}
#Retrieve DEG genes in HLA high
HLA_DR_genes <- grep("^HLA-DR", rownames(combined_dds), value = T)
HLA_DP_genes <- grep("^HLA-DP", rownames(combined_dds), value = T)
HLA_DQ_genes <- grep("^HLA-DQ", rownames(combined_dds), value = T)
HLAII_genes <- unique(c("CIITA", HLA_DR_genes, HLA_DP_genes, HLA_DQ_genes))

combined_results_filter_less_strict <- as.data.frame(results(combined_dds, name = "KNN_cluster_HLA_high_vs_HLA_low",cooksCutoff = T)) %>% 
  dplyr::filter(padj < 0.01 & baseMean >30 & abs(log2FoldChange) > 1.5)
combined_results_filter_less_strict$gene <- rownames(combined_results_filter_less_strict)


immunological_dataset <- read.table("../public/immunological_gene_lists.txt", header = F, sep = "\t",fill = T)
immunological_dataset <- subset(immunological_dataset, V3 == "GO")

# ensembl <- useEnsembl("ensembl", dataset = "hsapiens_gene_ensembl")
load("../controlled_data/ensembl.RData") #saved it to ensure it works when ensembl is down
# 2. Get all genes annotated to immune GO terms
go_terms <- immunological_dataset$V2

immune_anno <- getBM(
  attributes = c("hgnc_symbol", "go_id"),  # or "hgnc_symbol"
  filters    = "go",                              # correct filter name
  values     = go_terms,
  mart       = ensembl
)


immune_genes <- unique(immune_anno$hgnc_symbol)

# 3. Filter your own list
my_genes       <- combined_results_filter_less_strict$gene
filtered_genes <- setdiff(my_genes, immune_genes)

combined_results_filter_no_immune <- subset(combined_results_filter_less_strict, gene %in% filtered_genes)

# top 15 up
up15 <- combined_results_filter_no_immune %>%
  dplyr::filter(baseMean > 800 & log2FoldChange > 2.5) %>%
 top_n(15, wt = log2FoldChange)

# top 15 *down* (note the negative n)
down15 <- combined_results_filter_no_immune %>%
  dplyr::filter(baseMean > 800 & log2FoldChange < -2.5) %>%
  top_n(-30, wt = log2FoldChange)

# combine
gene_signature <- dplyr::bind_rows(up15, down15) %>%
  dplyr::arrange(desc(log2FoldChange))
gene_signature

#manually pruning immune-genes
# genes_to_remove <- c("CCL2", "IL24", "CIITA", "CXCL8", "IL1B")
genes_to_remove <- c("CIITA")
genes_to_remove_i <- which(rownames(gene_signature) %in% genes_to_remove)
if (length(genes_to_remove_i) == 0){
  gene_signature_manually_curated <- gene_signature
} else {
  gene_signature_manually_curated <- gene_signature[-genes_to_remove_i,]
}


# define your lists
up_list   <- rownames(subset(gene_signature_manually_curated, log2FoldChange > 0))
down_list <- rownames(subset(gene_signature_manually_curated, log2FoldChange < 0))
gene_signature_manually_curated
cat(paste0(gene_signature_manually_curated$gene," ", gene_signature_manually_curated$log2FoldChange ))
length(gene_signature_manually_curated$gene)
```

## Comparison with other signatures

What is the level of overlap of the signature with Kim et al, JCI 2021? How well does it predict HLA-II expression in the TCL data?

```{r, fig.width=7, fig.height=7}
dediff_kim <- read.xlsx("E:/PhD_projects/MHCII_project/TCL_analyses/Kim_dediff_signature.xlsx")

kim_genes <- unique(na.omit(dediff_kim$DediffGenes))
dediff_genes <- unique(na.omit(subset(combined_results_filter_less_strict,log2FoldChange>0)[["gene"]]))
signature <- unique(na.omit(subset(gene_signature_manually_curated,log2FoldChange>0)[["gene"]]))
reactome_HLA_II <- msigdbr(db_species = "HS", collection = "C2", subcollection = "CP:REACTOME") %>%
  dplyr::filter(gs_name == "REACTOME_MHC_CLASS_II_ANTIGEN_PRESENTATION") %>% dplyr::pull(gene_symbol) %>% unique()

gene_sets <- list(
  Kim_dediff = kim_genes,
  HLA_II_CONST_signature_raw = dediff_genes,
  HLA_II_CONST_signature_clean = signature,
  reactome_HLA_II = reactome_HLA_II
)

ggVennDiagram::ggVennDiagram(gene_sets, label = "count") +
  scale_fill_gradient(low = "white", high = "red") +
  theme_void(base_size = 20)

combined_vst <- vst(combined_dds, blind = TRUE)
expr <- assay(combined_vst)
expr_mat <- as.matrix(expr)
mode(expr_mat) <- "numeric"

# de-duplicate genes if needed (keep the most variable probe/gene)
if (any(duplicated(rownames(expr_mat)))) {
  expr_mat <- expr_mat[!duplicated(rownames(expr_mat)), ]
}

gsparam_kim <- ssgseaParam(exprData = expr_mat, geneSets = gene_sets)
ssgsea_scores_kim <- GSVA::gsva(param = gsparam_kim,verbose = TRUE)
ssgsea_z_kim <- t(scale(t(ssgsea_scores_kim)))

annots <- combined_dds$KNN_cluster
names(annots) <- colnames(combined_dds)
ann_colors <- list(
  HLA_status = c(
    HLA_low        = "#0384fc",
    HLA_mid_low         = "darkblue",
    HLA_mid_high= "#8b0000",
    HLA_high   = "red"
  )
)
levels4 <- c("HLA_high", "HLA_mid_high", "HLA_mid_low", "HLA_low") 
annot_df_kim <- data.frame(
  HLA_status = factor(annots[colnames(ssgsea_z_kim)], levels = levels4)
)

CIITA_counts <- assay(vst(combined_dds))["CIITA",]
IFN_dediff_score <- ssgsea_z_kim[ "Kim_dediff",]
IFN_dediff_score <- IFN_dediff_score[names(CIITA_counts)]
HLA_II_CONST_signature_raw_score <- ssgsea_z_kim[ "HLA_II_CONST_signature_raw",]
HLA_II_CONST_signature_raw_score <- HLA_II_CONST_signature_raw_score[names(CIITA_counts)]
HLA_II_CONST_signature_clean_score <- ssgsea_z_kim[ "HLA_II_CONST_signature_clean",]
HLA_II_CONST_signature_clean_score <- HLA_II_CONST_signature_clean_score[names(CIITA_counts)]
reactome_HLA_II_score <- ssgsea_z_kim[ "reactome_HLA_II",]
reactome_HLA_II_score <- reactome_HLA_II_score[names(CIITA_counts)]



scatterdata <- rbind(CIITA_counts, IFN_dediff_score, HLA_II_CONST_signature_raw_score, HLA_II_CONST_signature_clean_score, reactome_HLA_II_score)
scatterdata <- as.data.frame(t(scatterdata))
scatterdata$Sample <- rownames(scatterdata)
scatterdata<- scatterdata %>% left_join(as.data.frame(colData(combined_dds)), by = "Sample")


IFN_dediff_plot <- ggscatter(
  scatterdata,
  x = "CIITA_counts",
  y = "IFN_dediff_score",
  shape = "Dataset",
  fill = "KNN_cluster",
  color = "KNN_cluster",
  merge = TRUE,
  cor.coef = TRUE,
  title = "IFN dediff score - CIITA expression"
) +
  geom_smooth(
    data = scatterdata,
    mapping = aes(
      x = CIITA_counts,
      y = IFN_dediff_score,
      group = 1
    ),
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    inherit.aes = FALSE,
    color = "black",
    fill = "grey70"
  ) +
  scale_color_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  )) +
  scale_fill_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  ))
IFN_dediff_plot

HLA_II_CONST_clean_plot <- ggscatter(
  scatterdata,
  x = "CIITA_counts",
  y = "HLA_II_CONST_signature_clean_score",
  shape = "Dataset",
  fill = "KNN_cluster",
  color = "KNN_cluster",
  merge = TRUE,
  cor.coef = TRUE,
  title = "HLA-II^CONST+ dediff score - CIITA expression"
) +
  geom_smooth(
    data = scatterdata,
    mapping = aes(
      x = CIITA_counts,
      y = HLA_II_CONST_signature_clean_score,
      group = 1
    ),
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    inherit.aes = FALSE,
    color = "black",
    fill = "grey70"
  ) +
  scale_color_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  )) +
  scale_fill_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  ))
HLA_II_CONST_clean_plot

HLA_II_CONST_raw_plot <- ggscatter(
  scatterdata,
  x = "CIITA_counts",
  y = "HLA_II_CONST_signature_raw_score",
  shape = "Dataset",
  fill = "KNN_cluster",
  color = "KNN_cluster",
  merge = TRUE,
  cor.coef = TRUE,
  title = "HLA-II^CONST+ dediff score raw - CIITA expression"
) +
  geom_smooth(
    data = scatterdata,
    mapping = aes(
      x = CIITA_counts,
      y = HLA_II_CONST_signature_raw_score,
      group = 1
    ),
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    inherit.aes = FALSE,
    color = "black",
    fill = "grey70"
  ) +
  scale_color_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  )) +
  scale_fill_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  ))
HLA_II_CONST_raw_plot

reactome_plot <- ggscatter(
  scatterdata,
  x = "CIITA_counts",
  y = "reactome_HLA_II_score",
  shape = "Dataset",
  fill = "KNN_cluster",
  color = "KNN_cluster",
  merge = TRUE,
  cor.coef = TRUE,
  title = "HLA-II reactome - CIITA expression"
) +
  geom_smooth(
    data = scatterdata,
    mapping = aes(
      x = CIITA_counts,
      y = reactome_HLA_II_score,
      group = 1
    ),
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    inherit.aes = FALSE,
    color = "black",
    fill = "grey70"
  ) +
  scale_color_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  )) +
  scale_fill_manual(values = c(
    HLA_low = "#0384fc",
    HLA_mid_low = "darkblue",
    HLA_mid_high = "#8b0000",
    HLA_high = "red"
  ))
reactome_plot
```

# VALIDATION

## Get in House biopsy data
```{r}
source("TIL_data_preparation.R")

stopifnot(is.matrix(TIL_counts))
sym <- rownames(TIL_counts)

# 1) biomaRt: SYMBOL -> Ensembl + transcript lengths
# ensembl <- useEnsembl(
#   biomart = "genes",
#   dataset = "hsapiens_gene_ensembl"
# )
annot <- getBM(
  attributes = c("hgnc_symbol","ensembl_gene_id","gene_biotype",
                 "ensembl_transcript_id","transcript_length","transcript_is_canonical"),
  filters    = "hgnc_symbol",
  values     = unique(sym),
  mart       = ensembl
)

# Keep only rows with symbol present in your matrix
annot <- annot %>% dplyr::filter(hgnc_symbol %in% sym)

# 2) Prefer protein-coding; prefer canonical transcripts if present
annot_filt <- annot %>%
  dplyr::filter(gene_biotype == "protein_coding") %>%
  mutate(is_canonical = transcript_is_canonical == 1)

# 3) Compute one length per SYMBOL (in BASES)
# First, within each SYMBOL pick canonical transcripts if any; else use all.
length_by_symbol <- annot_filt %>%
  group_by(hgnc_symbol) %>%
  dplyr::mutate(use_canonical = any(is_canonical)) %>%
  dplyr::filter(if (dplyr::first(use_canonical)) is_canonical else TRUE) %>%
  summarise(gene_length_bases = median(transcript_length, na.rm = TRUE), .groups = "drop")

# Fallback: for symbols that dropped out (no protein_coding or NA), use any biotype
missing_syms <- setdiff(sym, length_by_symbol$hgnc_symbol)
if (length(missing_syms) > 0) {
  fallback <- annot %>%
    dplyr::filter(hgnc_symbol %in% missing_syms) %>%
    dplyr::group_by(hgnc_symbol) %>%
    dplyr::summarise(gene_length_bases = median(transcript_length, na.rm = TRUE), .groups = "drop")
  length_by_symbol <- bind_rows(length_by_symbol, fallback) %>%
    distinct(hgnc_symbol, .keep_all = TRUE)
}

# 4) Align lengths to counts (BASES; do NOT divide by 1000)
gene_len <- length_by_symbol$gene_length_bases[match(sym, length_by_symbol$hgnc_symbol)]

if (any(duplicated(sym))) {
  message("Duplicated symbols detected; collapsing by sum.")
  TIL_counts <- rowsum(TIL_counts, group = sym, reorder = FALSE)
  sym <- rownames(TIL_counts)
  gene_len <- length_by_symbol$gene_length_bases[match(sym, length_by_symbol$hgnc_symbol)]
}

# Warn about missing lengths
na_len <- is.na(gene_len)
if (any(na_len)) {
  message(sprintf("Dropping %d genes with missing lengths before RPKM.", sum(na_len)))
}

# 5) Create DGEList with lengths (in BASES)
dge <- DGEList(counts = TIL_counts[!na_len, , drop = FALSE],
               genes  = data.frame(Symbol = sym[!na_len],
                                   Length = gene_len[!na_len],
                                   row.names = sym[!na_len]))

# 6) TMM normalization
dge <- calcNormFactors(dge)

# 7) Compute RPKM/FPKM
# edgeR::rpkm expects 'gene.length' in BASES and divides by 1000 internally.
rpkm_mat <- rpkm(dge, gene.length = dge$genes$Length, normalized.lib.sizes = TRUE)

# TPM not used but code left for future changes
# TPM = scale RPK to 1e6
# RPK = counts / (length_kb); length_kb = length_bases / 1000
length_kb <- dge$genes$Length / 1000
rpk <- sweep(dge$counts, 1, length_kb, "/")
tpm_mat <- sweep(rpk, 2, colSums(rpk), "/") * 1e6

rank_data_til <- rankGenes(rpkm_mat)
```

## Compare biopsy signature score with TCL HLA-II expression

```{r concordance among signature and flow data from tcls}
tcl_data <- read.xlsx("../controlled_data/TIL/MHC_expression_TCLs.xlsx",
                      sheet = "Final Data")

scores_fortcl <- simpleScore(rank_data_til, upSet = up_list, downSet = down_list, centerScore = TRUE)
scores_fortcl$sample <- rownames(scores_fortcl)

validation_signature <- tcl_data %>%
  mutate(sample = matching_col) %>%
  inner_join(dplyr::select(scores_fortcl, sample, sig = TotalScore), by = "sample") %>%
  transmute(
    sample,
    fc       = as.numeric(FC.Baseline.to.Isotype),
    fc_log   = log10(pmax(fc, 1e-4)),              # log-transform; protects from 0
    sig      = as.numeric(sig),
    fc_ifn   = as.numeric(FC.IFN.to.isotype),
    fc_ifn_log   = log10(pmax(fc_ifn, 1e-4)),
    reac_base = as.numeric(reactivity_baseline_YTIL),
    reac_IFNg = as.numeric(reactivity_IFNg_yTIL)
  )

## 1) HLA-II FC vs isotype
cc_fc <- complete.cases(validation_signature$sig, validation_signature$fc)
ct_fc <- cor.test(validation_signature$sig[cc_fc],
                  validation_signature$fc_log[cc_fc],
                  method = "spearman", exact = FALSE)

ggplot(validation_signature, aes(sig, fc_log)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy signature score",
    y = "HLA-II FC vs isotype",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_fc$estimate, ct_fc$p.value),
    caption = paste0("N = ", sum(cc_fc))
  ) +
  theme_classic()

## 2) CD4+ TIL reactivity (untreated)
cc_reac <- complete.cases(validation_signature$sig, validation_signature$reac_base)
ct_reac <- cor.test(validation_signature$sig[cc_reac],
                    validation_signature$reac_base[cc_reac],
                    method = "spearman", exact = FALSE)

ggplot(validation_signature, aes(sig, reac_base)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy signature score",
    y = "CD4+ TIL Reactivity (untreated)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_reac$estimate, ct_reac$p.value),
    caption = paste0("N = ", sum(cc_reac))
  ) +
  theme_classic()

## 3) CD4+ TIL reactivity (IFN)
cc_reac_ifn <- complete.cases(validation_signature$sig, validation_signature$reac_IFNg)
ct_reac_ifn <- cor.test(validation_signature$sig[cc_reac_ifn],
                        validation_signature$reac_IFNg[cc_reac_ifn],
                        method = "spearman", exact = FALSE)

ggplot(validation_signature, aes(sig, reac_IFNg)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy signature score",
    y = "CD4 TIL reactivity (IFN)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_reac_ifn$estimate, ct_reac_ifn$p.value),
    caption = paste0("N = ", sum(cc_reac_ifn))
  ) +
  theme_classic()

## 4) HLA-II FC vs isotype (IFN condition)
cc_ifn <- complete.cases(validation_signature$sig, validation_signature$fc_ifn_log)
ct_ifn <- cor.test(validation_signature$sig[cc_ifn],
                   validation_signature$fc_ifn_log[cc_ifn],
                   method = "spearman", exact = FALSE)

ggplot(validation_signature, aes(sig, fc_ifn_log)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy signature score",
    y = "HLA-II FC vs isotype (IFN)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_ifn$estimate, ct_ifn$p.value),
    caption = paste0("N = ", sum(cc_ifn))
  ) +
  theme_classic()

## this is the result using class II genes only
scores_fortcl_HLA <- simpleScore(rank_data_til, upSet = HLAII_genes, centerScore = TRUE)
scores_fortcl_HLA$sample <- rownames(scores_fortcl)

validation_signature_HLA <- tcl_data %>%
  mutate(sample = matching_col) %>%
  inner_join(dplyr::select(scores_fortcl_HLA, sample, sig = TotalScore), by = "sample") %>%
  transmute(
    sample,
    fc       = as.numeric(FC.Baseline.to.Isotype),
    fc_log   = log10(pmax(fc, 1e-4)),              # log-transform; protects from 0
    sig      = as.numeric(sig),
    fc_ifn   = as.numeric(FC.IFN.to.isotype),
    fc_ifn_log   = log10(pmax(fc_ifn, 1e-4)),
    reac_base = as.numeric(reactivity_baseline_YTIL),
    reac_IFNg = as.numeric(reactivity_IFNg_yTIL)
  )

## 1) HLA-II FC vs isotype
cc_fc_hla <- complete.cases(validation_signature_HLA$sig, validation_signature_HLA$fc_log)
ct_fc_hla <- cor.test(validation_signature_HLA$sig[cc_fc_hla],
                  validation_signature_HLA$fc_log[cc_fc_hla],
                  method = "spearman", exact = FALSE)

ggplot(validation_signature_HLA, aes(sig, fc_log)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy HLA-II score",
    y = "HLA-II FC vs isotype",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_fc_hla$estimate, ct_fc_hla$p.value),
    caption = paste0("N = ", sum(cc_fc))
  ) +
  theme_classic()

## 2) CD4+ TIL reactivity (untreated)
cc_reac_hla <- complete.cases(validation_signature_HLA$sig, validation_signature_HLA$reac_base)
ct_reac <- cor.test(validation_signature_HLA$sig[cc_reac_hla],
                    validation_signature_HLA$reac_base[cc_reac_hla],
                    method = "spearman", exact = FALSE)

ggplot(validation_signature_HLA, aes(sig, reac_base)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy HLA-II score",
    y = "CD4+ TIL Reactivity (untreated)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_reac$estimate, ct_reac$p.value),
    caption = paste0("N = ", sum(cc_reac))
  ) +
  theme_classic()

## 3) CD4+ TIL reactivity (IFN)
cc_reac_ifn_hla <- complete.cases(validation_signature_HLA$sig, validation_signature_HLA$reac_IFNg)
ct_reac_ifn <- cor.test(validation_signature_HLA$sig[cc_reac_ifn_hla],
                        validation_signature_HLA$reac_IFNg[cc_reac_ifn_hla],
                        method = "spearman", exact = FALSE)

ggplot(validation_signature_HLA, aes(sig, reac_IFNg)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy HLA-II score",
    y = "CD4 TIL reactivity (IFN)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_reac_ifn$estimate, ct_reac_ifn$p.value),
    caption = paste0("N = ", sum(cc_reac_ifn))
  ) +
  theme_classic()

## 4) HLA-II FC vs isotype (IFN condition)
cc_ifn_hla <- complete.cases(validation_signature_HLA$sig, validation_signature_HLA$fc_ifn_log)
ct_ifn_hla <- cor.test(validation_signature_HLA$sig[cc_ifn_hla],
                   validation_signature_HLA$fc_ifn_log[cc_ifn_hla],
                   method = "spearman", exact = FALSE)

ggplot(validation_signature_HLA, aes(sig, fc_ifn_log)) +
  geom_point(size = 2, alpha = .8) +
  geom_smooth(method = "glm", se = FALSE) +
  labs(
    x = "Biopsy HLA-II score",
    y = "HLA-II FC vs isotype (IFN)",
    title   = sprintf("Spearman rho = %.2f, p = %.3g", ct_ifn_hla$estimate, ct_ifn_hla$p.value),
    caption = paste0("N = ", sum(cc_ifn))
  ) +
  theme_classic()
```

# SURVIVAL ANALYSIS

# GDC data preparation
```{r GDC download}
# 1. Query and download FPKM counts + clinical
# query <- GDCquery(project = "TCGA-SKCM",
#                   data.category = "Transcriptome Profiling",
#                   data.type     = "Gene Expression Quantification",
#                   workflow.type = "STAR - Counts")

# GDCdownload(query, directory = "../controlled_data/GDC_data/")
# dataSE_sckm <- GDCprepare(query, directory = "../../GDCdata/")

load("../controlled_data/GDCdata/TCGA-SKCM/dataSE_sckm.RData")
expr_skcm  <- assay(dataSE_sckm)               # sample × gene FPKM
clin_skcm  <- colData(dataSE_sckm)  
counts_mel = expr_skcm

clinical_data <- read.xlsx(xlsxFile ="../public/Survival_data_TCGA_upd.xlsx", rowNames = T)
clin_final <- as.data.frame(merge(x = clinical_data,y = clin_skcm, by.x = "bcr_patient_barcode", by.y = "patient"))

rownames(counts_mel) <- sub("\\..*$", "", rownames(counts_mel))
# 4) Retrieve gene lengths (median transcript length per gene) from Ensembl
# ensembl <- useEnsembl("ensembl", dataset="hsapiens_gene_ensembl")
geneAnnot <- getBM(
  attributes = c("ensembl_gene_id", "transcript_length"),
  filters    = "ensembl_gene_id",
  values     = rownames(counts_mel),
  mart       = ensembl
)
# compute median transcript length
library(dplyr)
geneLengths <- geneAnnot %>%
  dplyr::group_by(ensembl_gene_id) %>%
  dplyr::summarise(median_length = median(transcript_length)) %>%
  dplyr::ungroup()


# match lengths to your count matrix
length.kb <- geneLengths$median_length[match(rownames(counts_mel),
                                             geneLengths$ensembl_gene_id)] / 1000

# 5) Create DGEList and calculate normalization factors
dge <- DGEList(counts = counts_mel, genes = data.frame(Length = length.kb))
dge <- calcNormFactors(dge)  # TMM normalization

# 6) Compute FPKM
fpkm.mat_skcm <- rpkm(dge,
                 gene.length        = dge$genes$Length,
                 normalized.lib.sizes = TRUE)

library(org.Hs.eg.db)
library(AnnotationDbi)


# 2. Map IDs
hugo <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(fpkm.mat_skcm),
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)
```

# Administrative censoring at 15 years
```{r}
#administrative censoring at 15 years
clin_final$DSS.years <- clin_final$DSS.time/365
clin_final$DSS.years_censored   <- pmin(clin_final$DSS.years, 15)
clin_final$DSS_censored <- ifelse(clin_final$DSS.years > 15, 0, clin_final$DSS)
clin_final$OS.years <- clin_final$OS.time/365
clin_final$OS.years_censored   <- pmin(clin_final$OS.years, 15)
clin_final$OS_censored <- ifelse(clin_final$OS.years > 15, 0, clin_final$OS)
clin_final$PFI.years <- clin_final$PFI.time/365
clin_final$PFI.years_censored   <- pmin(clin_final$PFI.years, 15)
clin_final$PFI_censored <- ifelse(clin_final$PFI.years > 15, 0, clin_final$PFI)

cat(paste0("Administrative censoring: ", round((length(which(clin_final$DSS.years > 15))/length(clin_final$DSS.years))*100), "%"))
```



# Univariate analysis on TCGA-SKCM
```{r CONST+ score and univariate skcm, fig.width=15, fig.height=8}
#Subset out NAs and rename the exp matrix
keep <- !is.na(hugo)
expr_skcm <- fpkm.mat_skcm[keep, , drop=FALSE]
rownames(expr_skcm) <- hugo[keep]

#get clinical data
md <- read.table("../public/md_SKCM.txt", header = T, sep = "\t")
# define your lists
up_list   <- rownames(subset(gene_signature_manually_curated, log2FoldChange > 0))
down_list <- rownames(subset(gene_signature_manually_curated, log2FoldChange < 0))

rank_data <- rankGenes(expr_skcm)

# compute simple singscore
scores <- simpleScore(rank_data,
                      upSet   = up_list,
                      downSet = down_list)
rownames(scores) <- gsub(x= rownames(scores), pattern = "\\.", replacement = "-")
scores$sample <- rownames(scores)
scores <- subset(scores, sample %in% rownames(md))

clin_final <- subset(clin_final, barcode %in% scores$sample)
clin_final$sample <- clin_final$barcode
scores <- scores[order(scores$sample, clin_final$sample),]

# assuming rownames(md) are your sample IDs:
clin_final$signature_score <- scores$TotalScore[
  match(clin_final$sample, scores$sample)
]

# Check they all matched:
stopifnot(!any(is.na(clin_final$signature_score)))

clin_final <- clin_final %>% as.data.frame() %>%
  dplyr::mutate(group = if_else(signature_score >= 
                           quantile(signature_score, 0.75),
                         "High","Low"))
clin_final$group <- relevel(as.factor(clin_final$group), ref = "Low")
library(survival)
library(survminer)

#Fit the survival model with a continuous z-score as a sensitivity analysis
clin_final$score_z <- as.numeric(scale(clin_final$signature_score))

fit_os_cont <- coxph(
  Surv(OS.years_censored, OS_censored) ~ score_z,
  data = clin_final
)
summary(fit_os_cont)

fit_pfs_cont <- coxph(
  Surv(PFI.years_censored, PFI_censored) ~ score_z,
  data = clin_final
)
summary(fit_pfs_cont)

# 4) Fit the survival model in one go, using the data frame

fit_dss <- survfit(
  Surv(DSS.years_censored, DSS_censored) ~ group,
  data = clin_final
)

#proportional hazard model
cat("PH test p value: ", cox.zph(coxph(Surv(DSS.years_censored, DSS_censored) ~ group, data = clin_final))[["table"]]["GLOBAL","p"])


names(fit_dss$strata) <- gsub("group=","", names(fit_dss$strata))
s_cox <- coxph(Surv(DSS.years_censored, DSS_censored) ~ group, data = clin_final)
s_cox   <- summary(s_cox)
hr     <- s_cox$coefficients[, "exp(coef)"]
ci_lo  <- s_cox$conf.int[, "lower .95"]
ci_hi  <- s_cox$conf.int[, "upper .95"]
pval   <- s_cox$coefficients[, "Pr(>|z|)"]
hr_lab <- sprintf("HR=%.2f (95%% CI %.2f–%.2f)", hr, ci_lo, ci_hi)
p_lab  <- paste0("p = ", sprintf("%.4f", pval))

# 5a) plot DSS
ggsurvfit(fit_dss, linewidth=0.8)+
    add_censor_mark() +
    add_risktable(
        size            = 7,
        theme           = theme_risktable_default(plot.title.size  = 15),
        risktable_stats = "{n.risk} ({cum.event})",
        stats_label     = "Number at risk",risktable_height = 0.2
    ) +add_risktable_strata_symbol(symbol = "•", size = 15)+
    labs(
        x = "Years",
        y = paste0("Disease specific survival", " (%)")
    ) +
  scale_ggsurvfit(x_scales = list(limits = c(0,15), breaks=seq(0,15, by=3)))+
    scale_y_continuous(limits = c(0, 1),
                       labels = scales::percent_format(accuracy = 1),
                       expand = c(0, 0)) +
    theme_classic() +
    theme(
        plot.title       = element_text(hjust = 0.5, size = 20),
        plot.subtitle    = element_text(hjust = 0.5, size = 18),
        axis.title.x     = element_text(size = 20),
        axis.title.y     = element_text(size = 20, margin = margin(t = 0, r = 0, b = 0, l = 0)),
        axis.text.x      = element_text(size = 15),
        axis.text.y      = element_text(size = 15),
        legend.position  = "top",
        legend.direction = "vertical",
        legend.text      = element_text(size = 25),
        legend.key.size  = unit(30, "bigpts"),
        legend.title     = element_blank(),
        plot.margin      = unit(c(0,0.2,0,0), 'lines')
    )+scale_color_manual(values = c("High" = "red", "Low" = "blue"))+
  ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.2,
        label = hr_lab,
        size  = 10,
        hjust = 0
      ) +
      ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.1,
        label = p_lab,
        size  = 10,
        hjust = 0
      )

fit_os <- survfit(
  Surv(OS.years_censored, OS_censored) ~ group,
  data = clin_final
)  

fit_pfs <- survfit(
  Surv(PFI.years_censored, PFI_censored) ~ group,
  data = clin_final
)

# 5b) Plot
ggsurvplot(
  fit_os,
  data       = clin_final,
  pval       = TRUE,
  risk.table = TRUE
)

# 5c) Plot PFS
ggsurvplot(
  fit_pfs,
  data       = clin_final,
  pval       = TRUE,
  risk.table = TRUE
)


```

## Subgroup analysis on the few ICI treated patients in TCGA data

```{r, fig.width=15, fig.height=8}
library(data.table)
clin_ici_tcga <- fread("E:/PhD_projects/MHCII_project/TCGA_SKCM/clinical_data_ICI_patients_SKCM/clinical.tsv")
ici_treated_tcga_skcm <- unique(clin_ici_tcga$cases.submitter_id)
clin_final <- clin_final %>% mutate(ICI= case_when(bcr_patient_barcode %in% ici_treated_tcga_skcm ~ "Yes",
                                                             !(bcr_patient_barcode %in% ici_treated_tcga_skcm) ~ "No")) 
table(clin_final$ICI)

fit_cox_ICI <- survfit(
  Surv(DSS.years_censored, DSS_censored) ~ group,
  data = dplyr::filter(clin_final, ICI == "Yes")
)

names(fit_cox_ICI$strata) <- gsub("group=","", names(fit_cox_ICI$strata))
s_cox_ICI <- coxph(Surv(DSS.years_censored, DSS_censored) ~ group, data = dplyr::filter(clin_final, ICI == "Yes"))
s_cox_ICI   <- summary(s_cox_ICI)
hr     <- s_cox_ICI$coefficients[, "exp(coef)"]
ci_lo  <- s_cox_ICI$conf.int[, "lower .95"]
ci_hi  <- s_cox_ICI$conf.int[, "upper .95"]
pval   <- s_cox_ICI$coefficients[, "Pr(>|z|)"]
hr_lab <- sprintf("HR=%.2f (95%% CI %.2f–%.2f)", hr, ci_lo, ci_hi)
p_lab  <- paste0("p = ", sprintf("%.4f", pval))

# plot dss
ggsurvfit(fit_cox_ICI, linewidth=0.8)+
    add_censor_mark() +
    add_risktable(
        size            = 7,
        theme           = theme_risktable_default(plot.title.size  = 15),
        risktable_stats = "{n.risk} ({cum.event})",
        stats_label     = "Number at risk",risktable_height = 0.2
    ) +add_risktable_strata_symbol(symbol = "•", size = 15)+
    labs(
        x = "Years",
        y = paste0("Disease specific survival", " (%)")
    ) +
  scale_ggsurvfit(x_scales = list(limits = c(0,15), breaks=seq(0,15, by=3)))+
    scale_y_continuous(limits = c(0, 1),
                       labels = scales::percent_format(accuracy = 1),
                       expand = c(0, 0)) +
    theme_classic() +
    theme(
        plot.title       = element_text(hjust = 0.5, size = 20),
        plot.subtitle    = element_text(hjust = 0.5, size = 18),
        axis.title.x     = element_text(size = 20),
        axis.title.y     = element_text(size = 20, margin = margin(t = 0, r = 0, b = 0, l = 0)),
        axis.text.x      = element_text(size = 15),
        axis.text.y      = element_text(size = 15),
        legend.position  = "top",
        legend.direction = "vertical",
        legend.text      = element_text(size = 25),
        legend.key.size  = unit(30, "bigpts"),
        legend.title     = element_blank(),
        plot.margin      = unit(c(0,0.2,0,0), 'lines')
    )+scale_color_manual(values = c("High" = "red", "Low" = "blue"))+
  ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.2,
        label = hr_lab,
        size  = 10,
        hjust = 0
      ) +
      ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.1,
        label = p_lab,
        size  = 10,
        hjust = 0
      )
  

```

## Subgroup analysis on all the patients who ever received immunotherapy

```{r,fig.width=15, fig.height=8}
clin_immuno_tcga <- fread("../public/all_immunotherapies_SKCM.tsv")
immuno_treated_tcga_skcm <- unique(clin_immuno_tcga$`Case ID`)
clin_final <- clin_final %>% mutate(immunotherapy= case_when(bcr_patient_barcode %in% immuno_treated_tcga_skcm ~ "Yes",
                                                             !(bcr_patient_barcode %in% immuno_treated_tcga_skcm) ~ "No")) 
table(clin_final$immunotherapy)

fit_cox_immuno <- survfit(
  Surv(DSS.years_censored, DSS_censored) ~ group+ICI,
  data = dplyr::filter(clin_final, immunotherapy == "Yes")
)

names(fit_cox_immuno$strata) <- gsub("group=","", names(fit_cox_immuno$strata))
clin_final$ICI <- as.factor(clin_final$ICI)
s_cox_Immuno <- coxph(Surv(DSS.years_censored, DSS_censored) ~ group+ICI, data = dplyr::filter(clin_final, immunotherapy == "Yes"))
s_cox_Immuno   <- summary(s_cox_Immuno)
hr     <- s_cox_Immuno$coefficients[, "exp(coef)"]
ci_lo  <- s_cox_Immuno$conf.int[, "lower .95"]
ci_hi  <- s_cox_Immuno$conf.int[, "upper .95"]
pval   <- s_cox_Immuno$coefficients[, "Pr(>|z|)"]
hr_lab <- sprintf("HR=%.2f (95%% CI %.2f–%.2f)", hr, ci_lo, ci_hi)
p_lab  <- paste0("p = ", sprintf("%.4f", pval))

#plot dss
ggsurvfit(fit_cox_immuno, linewidth=0.8)+
    add_censor_mark() +
    add_risktable(
        size            = 7,
        theme           = theme_risktable_default(plot.title.size  = 15),
        risktable_stats = "{n.risk} ({cum.event})",
        stats_label     = "Number at risk",risktable_height = 0.2
    ) +add_risktable_strata_symbol(symbol = "•", size = 15)+
    labs(
        x = "Years",
        y = paste0("Disease specific survival", " (%)")
    ) +
  scale_ggsurvfit(x_scales = list(limits = c(0,15), breaks=seq(0,15, by=3)))+
    scale_y_continuous(limits = c(0, 1),
                       labels = scales::percent_format(accuracy = 1),
                       expand = c(0, 0)) +
    theme_classic() +
    theme(
        plot.title       = element_text(hjust = 0.5, size = 20),
        plot.subtitle    = element_text(hjust = 0.5, size = 18),
        axis.title.x     = element_text(size = 20),
        axis.title.y     = element_text(size = 20, margin = margin(t = 0, r = 0, b = 0, l = 0)),
        axis.text.x      = element_text(size = 15),
        axis.text.y      = element_text(size = 15),
        legend.position  = "top",
        legend.direction = "vertical",
        legend.text      = element_text(size = 25),
        legend.key.size  = unit(30, "bigpts"),
        legend.title     = element_blank(),
        plot.margin      = unit(c(0,0.2,0,0), 'lines')
    )+scale_color_manual(values = c("High, ICI=Yes" = "red", "Low, ICI=Yes" = "blue", "High, ICI=No " = "orange", "Low, ICI=No " = "darkgreen"))+
  ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.2,
        label = hr_lab,
        size  = 10,
        hjust = 0
      ) +
      ggplot2::annotate(
        "text",
        x     = 0,
        y     = 0.1,
        label = p_lab,
        size  = 10,
        hjust = 0
      )
  

```

## Prognostic vs predictive role of the signature
```{r}
#predictive for ici?
clin_tmp <- clin_final |>
    dplyr::mutate(
        ICI_bin = factor(ICI, levels = c("No","Yes")),   # ensure consistent ref
        sig_z   = as.numeric(scale(signature_score))                # z-score helps stability
    )

m_no_int <- coxph(
    Surv(DSS.years_censored, DSS_censored) ~ sig_z + immunotherapy,
    data = clin_tmp
)

m_int <- coxph(
    Surv(DSS.years_censored, DSS_censored) ~ sig_z * immunotherapy,
    data = clin_tmp
)

anova(m_no_int, m_int, test = "LRT")   # p-value here = evidence of predictiveness
summary(m_int)
cox.zph(m_int)
#predictive for all immunotherapies?
clin_tmp <- clin_final |>
    dplyr::mutate(
        ICI_bin = factor(ICI, levels = c("No","Yes")),   # ensure consistent ref
        sig_z   = as.numeric(scale(signature_score))                # z-score helps stability
    )

m_no_int <- coxph(
    Surv(DSS.years_censored, DSS_censored) ~ sig_z + immunotherapy,
    data = clin_tmp
)

m_int <- coxph(
    Surv(DSS.years_censored, DSS_censored) ~ sig_z * immunotherapy,
    data = clin_tmp
)

anova(m_no_int, m_int, test = "LRT")   # p-value here = evidence of predictiveness
summary(m_int)
cox.zph(m_int)
```

# Multivariate analysis on TCGA-SKCM


## Calculate Immune infiltration with ESTIMATE

```{r immune infiltration}
library(AnnotationDbi)
library(EnsDb.Hsapiens.v86)
library(IOBR)

# TPM from RPKM
tpm_mat_skcm <- t( t(fpkm.mat_skcm) / colSums(as.matrix(fpkm.mat_skcm),na.rm = T) * 1e6 )
tpm_log2_skcm <- log2(tpm_mat_skcm + 1)
#Conversion to HGNC gene names
ens <- rownames(tpm_log2_skcm)

sym <- mapIds(
  EnsDb.Hsapiens.v86,
  keys     = ens,
  keytype  = "GENEID",
  column   = "SYMBOL",
  multiVals = "first"
)

keep <- !is.na(sym) & sym != ""
tpm_log2_skcm <- tpm_log2_skcm[keep, , drop = FALSE]
rownames(tpm_log2_skcm) <- sym[keep]

dup <- duplicated(rownames(tpm_log2_skcm))
if (any(dup)) {
  o <- order(rownames(tpm_log2_skcm), rowMeans(tpm_log2_skcm), decreasing = TRUE)
  tpm_log2_skcm <- tpm_log2_skcm[o, ]
  tpm_log2_skcm <- tpm_log2_skcm[!duplicated(rownames(tpm_log2_skcm)), ]
}

#IOBR deconvolution
estimate_res_skcm <- IOBR::deconvo_tme(eset = tpm_log2_skcm, method = "estimate")
estimate_res_skcm_backup=estimate_res_skcm

#store immune score in a column in the clinical data
estimate_res_skcm <- subset(estimate_res_skcm, ID %in% clin_final$sample)
# assuming clin_final$sample are your sample IDs:
clin_final$immune_score <- estimate_res_skcm$ImmuneScore_estimate[
  match(clin_final$sample, estimate_res_skcm$ID)
]
```

### Sensitivity: Fit signature score as a continuous covariate

```{r, fig.width=10, fig.height=5}
#multiple cox
library(survival)
library(survminer)

clin_final$score_z <- scale(clin_final$signature_score)

univ_cox <- coxph(
  Surv(DSS.time, DSS) ~ score_z,
  data = clin_final
)
summary(univ_cox)


# 1) Recode age into <60 vs ≥60, and AJCC stage into 0‑2 / 3 / 4
clin_final2 <- clin_final %>%
  mutate(
    age_group = if_else(age_at_diagnosis/365 < 40, "<40", "≥40"),
    stage_group = case_when(
      ajcc_pathologic_tumor_stage %in% c("Stage 0", "I/II NOS",
                                   "Stage I","Stage IA","Stage IB",
                                   "Stage II","Stage IIA","Stage IIB","Stage IIC")
      ~ "0/I/II",
      ajcc_pathologic_tumor_stage %in% c("Stage III","Stage IIIA","Stage IIIB","Stage IIIC")
      ~ "III",
      ajcc_pathologic_tumor_stage == "Stage IV" 
      ~ "IV",
      TRUE ~ NA_character_
    ),
    Immune_infiltrate = if_else(condition = (immune_score >= quantile(immune_score, 0.75)),
                            "High","Low")
  ) %>%
  mutate(
    age_group   = factor(age_group,   levels=c("<40","≥40")),
    stage_group = factor(stage_group, levels=c("0/I/II","III","IV")),
    Immune_infiltrate = factor(Immune_infiltrate, levels = c("Low", "High")),
    Sex = as.factor(gender)
  ) %>%
  dplyr::filter(!is.na(age_group) & !is.na(stage_group))

multiv_cox <- coxph(
  Surv(DSS.time, DSS) ~ score_z + Immune_infiltrate + age_group + Sex + stage_group,
  data = clin_final2
)
summary(multiv_cox)
```


### Multivariate with dichotomized cohort

```{r, fig.width=10, fig.height=5}
clin_final2$group <- as.factor(clin_final2$group)
clin_final2$group <- relevel(clin_final2$group, ref = "Low")

cox_mod <- coxph(
  Surv(DSS.years_censored, DSS_censored) ~ 
    group     + 
    Immune_infiltrate +
    age_group   + 
    Sex      + 
    stage_group,
  data = clin_final2
)

summary(cox_mod)
ggforest(cox_mod, data=clin_final2, main = "DSS")

cox_mod_OS <- coxph(
  Surv(OS.years_censored, OS_censored) ~ 
    group     + 
    Immune_infiltrate +
    age_group   + 
    Sex      + 
    stage_group,
  data = clin_final2
)

summary(cox_mod_OS)
ggforest(cox_mod_OS, data=clin_final2, main = "OS")

cox_mod_PFS <- coxph(
  Surv(PFI.years_censored, PFI_censored) ~ 
    group     + 
    Immune_infiltrate +
    age_group   + 
    Sex      + 
    stage_group,
  data = clin_final2
)

summary(cox_mod_PFS)
ggforest(cox_mod_PFS, data=clin_final2, main = "PFS")
```

## Association with TMB
```{r}
# query <- GDCquery(project = "TCGA-SKCM",
#                   data.category = "Simple Nucleotide Variation",
#                   data.type     = "Masked Somatic Mutation"
#                   )
# GDCdownload(query = query, directory = "../controlled_data/GDCdata/")
# maf_sckm <- GDCprepare(query, directory = "../controlled_data/GDCdata")
load("../controlled_data/GDCdata/TCGA-SKCM/maf_skcm.RData")
total_muts <- as.data.frame(table(maf_sckm$Tumor_Sample_Barcode))

colnames(total_muts) <- c("Tumor_Sample_Barcode", "Mutations")

total_muts$bcr_patient_barcode  <- TCGAutils::TCGAbarcode(as.character(total_muts$Tumor_Sample_Barcode),participant = T)

clin_final3 <-merge(clin_final2, total_muts,by = "bcr_patient_barcode") 
clin_final3$TMB <- as.factor(ifelse(clin_final3$Mutations < quantile(clin_final3$Mutations, 0.5), "Low", "High"))
clin_final3$TMB <- relevel(clin_final3$TMB, ref = "Low")
cox_mod <- coxph(
  Surv(DSS.years_censored, DSS_censored) ~ 
    group     + 
    TMB +
    Immune_infiltrate +
    age_group   + 
    Sex      + 
    stage_group,
  data = clin_final3
)

# 3) Summarize and plot
summary(cox_mod)
ggforest(cox_mod, data=clin_final3)
```


## Baseline characteristics according to HLA group
```{r baseline char - hla group}
library(compareGroups)

df <- clin_final2 %>%
  transmute(group,
            is_primary = case_when(
              definition == "Primary solid Tumor" ~ "Primary",
              definition %in% c("not reported", NA) ~ NA_character_,
              definition == "Metastatic" ~ "Metastatic"
            )) %>%
  dplyr::filter(!is.na(is_primary), !is.na(group))

tab <- table(df$group, df$is_primary)
tab

# Fisher's exact (robust for small counts)
ft <- fisher.test(tab)
ft

# Also show proportions by group
prop_tbl <- prop.table(tab, margin = 1) %>% round(3)
prop_tbl

#GLM to make adjusted comparison
addmargins(tab)
prop.table(tab, 1)

# risk difference (absolute difference in proportions)
rd <- prop_tbl["High","Metastatic"] - prop_tbl["Low","Metastatic"]  # High - Low
rd  # ≈ -0.116  (≈ –11.6 percentage points)

# Phi coefficient (effect size for 2x2)
phi <- psych::phi(tab)  # needs psych package
phi

library(binom)
prop_df <- df %>%
  dplyr::count(group, is_primary) %>%
  pivot_wider(names_from = is_primary, values_from = n, values_fill = 0) %>%
  dplyr::rename(Metastatic = `Metastatic`, Primary = `Primary`) %>%
  mutate(n = Metastatic + Primary) %>%
  rowwise() %>%
  mutate(binom = list(binom.confint(x = Metastatic, n = n, methods = "wilson"))) %>%
  unnest(binom, names_sep = "_") %>%      # <-- avoids name collision
  transmute(group,
            p   = binom_mean,                   # from binom_*
            lwr = binom_lower,
            upr = binom_upper,
            n)

# y-position for the bracket (slightly above the tallest CI bar)
y_bracket <- max(prop_df$upr) * 1.08

# Build the p-value data frame for ggpubr
pdat <- data.frame(
  group1 = "Low",
  group2 = "High",
  y.position = y_bracket,
  p = ft$p.value,
  p.signif = symnum(ft$p.value, corr = FALSE, na = FALSE,
                    cutpoints = c(0, .001, .01, .05, .1, 1),
                    symbols = c("***","**","*",".","ns"))
)

ggplot(prop_df, aes(group, p, fill = group)) +
  geom_col(width = .6, color = "grey20") +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = .15, linewidth = .6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, max(prop_df$upr)*1.2)) +
  scale_fill_manual(values = c(Low = "#0384fc", High = "red")) +
  labs(title = "Proportion of Metastatic tumors by group",
       x = NULL, y = "Primary (proportion)") +
  theme_classic() + theme(legend.position = "none") +
  ggpubr::stat_pvalue_manual(pdat, label = "p.signif", tip.length = 0.01)


```

## Larger multivariate model (immune infiltration, CIITA, and single HLA-II isotypes)

```{r,  fig.width=10, fig.height=5}
# Test with Immune infiltration, CIITA, HLA-DRA, HLA-DP, HLA-DQ
###HLA-DR
grep("HLA-DR", rownames(rank_data), value=T)
# compute simple singscore
scores <- simpleScore(rank_data,
                      upSet   = unique(grep("HLA-DR", rownames(rank_data), value=T)))
rownames(scores) <- gsub(x= rownames(scores), pattern = "\\.", replacement = "-")
scores$sample <- rownames(scores)
scores <- subset(scores, sample %in% clin_final$sample)

clin_final$HLADR_score <- scores$TotalScore[
  match(clin_final$sample, scores$sample)
]
###HLA-DQ
scores <- simpleScore(rank_data,
                      upSet   = unique(grep("HLA-DQ", rownames(rank_data), value = T)))
rownames(scores) <- gsub(x= rownames(scores), pattern = "\\.", replacement = "-")
scores$sample <- rownames(scores)
scores <- subset(scores, sample %in% clin_final$sample)

clin_final$HLADQ_score <- scores$TotalScore[
  match(clin_final$sample, scores$sample)
]
###HLA-DP
scores <- simpleScore(rank_data,
                      upSet   = unique(grep("HLA-DP", rownames(rank_data), value = T)))
rownames(scores) <- gsub(x= rownames(scores), pattern = "\\.", replacement = "-")
scores$sample <- rownames(scores)
scores <- subset(scores, sample %in% clin_final$sample)

clin_final$HLADP_score <- scores$TotalScore[
  match(clin_final$sample, scores$sample)
]

###HLA-DP
scores <- simpleScore(rank_data,
                      upSet   = unique(grep("CIITA", rownames(rank_data), value = T)))
rownames(scores) <- gsub(x= rownames(scores), pattern = "\\.", replacement = "-")
scores$sample <- rownames(scores)
scores <- subset(scores, sample %in% clin_final$sample)

clin_final$CIITA_score <- scores$TotalScore[
  match(clin_final$sample, scores$sample)
]

clin_final <- clin_final %>%
  mutate(group_DR = if_else(HLADR_score >= 
                           quantile(HLADR_score, 0.75),
                         "High","Low"))
clin_final <- clin_final %>%
  mutate(group_DP = if_else(HLADP_score >= 
                           quantile(HLADP_score, 0.75),
                         "High","Low"))
clin_final <- clin_final %>%
  mutate(group_DQ = if_else(HLADQ_score >= 
                           quantile(HLADQ_score, 0.75),
                         "High","Low"))

clin_final <- clin_final %>%
  mutate(CIITA_group = if_else(CIITA_score >= 
                              quantile(CIITA_score, 0.75),
                            "High","Low"))
clin_final <- clin_final %>%
  mutate(Immune_infiltrate = if_else(immune_score >= 
                              quantile(immune_score, 0.75),
                            "High","Low"))

fit <- survfit(
  Surv(DSS.years_censored, DSS_censored) ~ Immune_infiltrate,
  data = clin_final
)

# Plot DSS for immune infiltration groups
ggsurvplot(
  fit,
  data       = clin_final,
  pval       = TRUE,
  risk.table = TRUE,
) + labs(
  title="Survival in Immune infiltration groups"
)

clin_final2 <- clin_final %>%
  mutate(
    Age_group = if_else(age_at_diagnosis/365 < 40, "<40", "≥40"),
    Stage_group = case_when(
      ajcc_pathologic_tumor_stage %in% c("Stage 0", "I/II NOS",
                                   "Stage I","Stage IA","Stage IB",
                                   "Stage II","Stage IIA","Stage IIB","Stage IIC")
      ~ "0/I/II",
      ajcc_pathologic_tumor_stage %in% c("Stage III","Stage IIIA","Stage IIIB","Stage IIIC")
      ~ "III",
      ajcc_pathologic_tumor_stage == "Stage IV" 
      ~ "IV",
      TRUE ~ NA_character_
    ),
    Immune_infiltrate = if_else(condition = (immune_score >= quantile(immune_score, 0.75)),
                            "High","Low")
  ) %>%
  mutate(
    Age_group   = factor(Age_group,   levels=c("<40","≥40")),
    Stage_group = factor(Stage_group, levels=c("0/I/II","III","IV")),
    Immune_infiltrate = factor(Immune_infiltrate, levels = c("Low", "High")),
    Sex = as.factor(gender)
  ) %>%
  dplyr::filter(!is.na(Age_group) & !is.na(Stage_group))

clin_final2$HLA_const <- as.factor(clin_final2$group)
clin_final2$HLA_const <- relevel(clin_final2$HLA_const, ref = "Low")
clin_final2$group_DR <- as.factor(clin_final2$group_DR)
clin_final2$group_DR <- relevel(clin_final2$group_DR, ref = "Low")
clin_final2$group_DP <- as.factor(clin_final2$group_DP)
clin_final2$group_DP <- relevel(clin_final2$group_DP, ref = "Low")
clin_final2$group_DQ <- as.factor(clin_final2$group_DQ)
clin_final2$group_DQ <- relevel(clin_final2$group_DQ, ref = "Low")
clin_final2$Immune_infiltrate <- as.factor(clin_final2$Immune_infiltrate)
clin_final2$Immune_infiltrate <- relevel(clin_final2$Immune_infiltrate, ref = "Low")
# clin_final2$Sex <- as.factor(clin_final2$Sex)
# clin_final2$Sex <- relevel(clin_final2$Sex, ref = "MALE")
clin_final2$CIITA_group <- as.factor(clin_final2$CIITA_group)
clin_final2$CIITA_group <- relevel(clin_final2$CIITA_group, ref = "Low")

# Fit the Cox model with the new categorical vars
cox_mod <- coxph(
  Surv(DSS.years_censored, DSS_censored) ~ 
    HLA_const     + 
    Immune_infiltrate +
    Age_group   + 
    Sex      + 
    Stage_group,
  data = clin_final2
)

# Summarize and plot
summary(cox_mod)
ggforest(cox_mod, data=clin_final2, main = "Multivariate Cox - HLACONST+ and Immune infiltrate")

# Check proportional‑hazards assumption
cox.zph(cox_mod)

# Fit the Cox model with the new categorical vars
cox_mod_CIITA <- coxph(
  Surv(DSS.years_censored, DSS_censored) ~ 
    HLA_const     + 
    Immune_infiltrate +
    CIITA_group +
    Age_group   + 
    Sex      + 
    Stage_group,
  data = clin_final2
)

# Summarize and plot
summary(cox_mod_CIITA)
ggforest(cox_mod_CIITA, data=clin_final2, main = "Multivariate Cox - Complete model")

cox_mod_complete <- coxph(
  Surv(DSS.years_censored, DSS_censored) ~ 
    HLA_const     + 
    Immune_infiltrate +
    CIITA_group +
    group_DR+
    group_DP+
    group_DQ+
    Age_group   + 
    Sex      + 
    Stage_group,
  data = clin_final2
)

# Summarize and plot
summary(cox_mod_complete)
ggforest(cox_mod_complete, data=clin_final2, main = "Multivariate Cox - Complete model")

# Check proportional‑hazards assumption
cox.zph(cox_mod_complete)

# show collinearity between immune infiltrate, CIITA and HLAs - simple correlations
vars <- c("score_z","immune_score","CIITA_score","HLADR_score","HLADP_score","HLADQ_score","age_at_diagnosis")
M <- clin_final[, vars]
colnames(M) <- c("Normalized HLA-II score", "Immune infiltrate", "CIITA expression", "HLA-DR expression", "HLA-DP expression", "HLA-DQ expression", "Age")
cors <- cor(M, use="pairwise.complete.obs", method="spearman")
cors

corrplot::corrplot(
  cors,
  method = "color",
  type = "upper",
  order = "hclust",
  addCoef.col = "black",
  tl.col = "black",
  col = colorRampPalette(c("navy", "white", "firebrick3"))(200),
  tl.srt = 45,
  diag = T
)


# collinearity in the design matrix
X <- model.matrix(cox_mod_complete)
X <- X[, colnames(X) != "(Intercept)", drop=FALSE]

# Remove columns with zero variance (can happen with sparse factor levels)
X <- X[, apply(X, 2, sd, na.rm=TRUE) > 0, drop=FALSE]

# Correlation matrix among columns
R <- cor(X, use="pairwise.complete.obs")

# VIF = diagonal of inverse correlation matrix
VIF <- diag(solve(R))
sort(VIF, decreasing = TRUE)
```

## Compare the different models (Exploratory, not included)

```{r}
library(survival)
library(broom)
library(dplyr)
library(ggplot2)
library(forcats)


# --- keep the same adjustments you were using
controls <- c("Age_group","Sex","Stage_group")

mk_mod <- function(rhs_vars) {
  rhs <- paste(c(rhs_vars, controls), collapse = " + ")
  coxph(as.formula(paste("Surv(DSS.years_censored, DSS_censored) ~", rhs)),
        data = clin_final2)
}

# --- your 4 models
mods <- list(
  "HLA+Immune infiltrate"        = mk_mod(c("HLA_const","Immune_infiltrate")),
  "HLA+CIITA"      = mk_mod(c("HLA_const","CIITA_group")),
  "Immune infiltrate+CIITA"      = mk_mod(c("Immune_infiltrate","CIITA_group")),
  "HLA+Immune infiltrate+CIITA"  = mk_mod(c("HLA_const","Immune_infiltrate","CIITA_group"))
)

# --- build a mapping of the term names used in the model for each variable

target_vars <- c("HLA_const","Immune_infiltrate","CIITA_group")
pretty_var  <- c(HLA_const="HLA Const+",
                 Immune_infiltrate="Immune infiltration",
                 CIITA_group="CIITA")

var_map <- bind_rows(lapply(target_vars, function(v){
  x <- clin_final2[[v]]
  if(!is.factor(x) || length(levels(x)) < 2)
    stop(sprintf("%s should be a 2+ level factor with first level = reference", v))
  ref  <- levels(x)[1]
  alt  <- levels(x)[2]              
  tibble(
    variable = v,
    term     = paste0(v, alt),      
    label    = sprintf("%s (%s vs %s)", pretty_var[[v]], alt, ref)
  )
}))

# --- tidy all models and keep only the 3 variables of interest
td <- bind_rows(lapply(names(mods), function(mn)
  broom::tidy(mods[[mn]], conf.int = TRUE, exponentiate = TRUE) |>
    mutate(model = mn)
)) |>
  inner_join(var_map, by = "term") |>
  transmute(
    var   = factor(label,
                   levels = sprintf("%s (%s vs %s)",
                                    pretty_var[target_vars],
                                    levels(clin_final2[[target_vars[1]]])[2],  
                                    levels(clin_final2[[target_vars[1]]])[1])),
    model = factor(model, levels = c("HLA+Immune infiltrate","HLA+CIITA","Immune infiltrate+CIITA","HLA+Immune infiltrate+CIITA")),
    HR    = estimate, lo = conf.low, hi = conf.high, p = p.value
  ) |>
  mutate(sig = p < 0.05)

# --- guard against zeros on log scale
td <- dplyr::filter(td, HR > 0, lo > 0, hi > 0)

# dynamic x-limits (log scale)
x_breaks <- c(0.25, 0.5, 1, 2, 4, 8)
x_min <- max(min(td$lo, na.rm = TRUE) * 0.8, min(x_breaks))
x_max <- max(td$hi, na.rm = TRUE) * 1.2

# --- horizontal comparison forest (one facet per variable)
ggplot(td, aes(y = model)) +
  facet_grid(var ~ ., switch = "y") +
  ggplot2::annotate("rect", xmin = x_min, xmax = 1, ymin = -Inf, ymax = Inf, alpha = 0.06) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_segment(aes(x = lo, xend = hi, yend = model, color = model), linewidth = 0.9) +
  geom_point(aes(x = HR, color = model, fill = sig, shape = sig), size = 2.8) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 21), guide = "none") +
  scale_fill_manual(values  = c(`TRUE` = "black", `FALSE` = NA), guide = "none") +
  labs(x = "Hazard ratio (log scale)", y = NULL,
       title = "Key covariates across models") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, hjust = 1),
        legend.position = "right")

mod_base <- mods[["HLA+Immune infiltrate"]]
mod_full <- mods[["HLA+Immune infiltrate+CIITA"]]

lr <- anova(mod_base, mod_full, test = "LRT")
lr_df <- as.data.frame(lr, stringsAsFactors = FALSE)

# Find columns safely
chi_col <- grep("Chisq", names(lr_df), value = TRUE)[1]
p_col   <- grep("^Pr\\(", names(lr_df), value = TRUE)[1]  # matches "Pr(>|Chi|)"
df_col  <- grep("^Df$",  names(lr_df), value = TRUE)[1]

chi   <- as.numeric(lr_df[[chi_col]][2])
p_LRT <- as.numeric(lr_df[[p_col]][2])
df_dx <- as.integer(lr_df[[df_col]][2])

ll_base <- as.numeric(logLik(mod_base))
ll_full <- as.numeric(logLik(mod_full))
df_base <- attr(logLik(mod_base), "df")
df_full <- attr(logLik(mod_full), "df")

chi   <- 2 * (ll_full - ll_base)
df_dx <- df_full - df_base
p_LRT <- pchisq(chi, df = df_dx, lower.tail = FALSE)

cap <- sprintf("Adding CIITA to HLA+Immune_infiltrate does not improve fit (LRT Δχ²=%.2f, df=%d, p=%s).",
               chi, df_dx,
               ifelse(p_LRT < 0.001, "<0.001", formatC(p_LRT, format="f", digits=3)))
cap
```

## Compare different HLA-II isotypes 

```{r}
# Terms to keep (adjust if your term names differ)
keep_terms <- c("group_DRHigh", "group_DPHigh", "group_DQHigh")

td <- broom::tidy(cox_mod_complete, exponentiate = TRUE, conf.int = TRUE) %>%
  dplyr::filter(term %in% keep_terms) %>%
  dplyr::mutate(
    isotype = case_when(
      term == "group_DRHigh" ~ "HLA-DR (High vs Low)",
      term == "group_DPHigh" ~ "HLA-DP (High vs Low)",
      term == "group_DQHigh" ~ "HLA-DQ (High vs Low)",
      TRUE ~ term
    ),
    # Optional: significance label
    p_label = case_when(
      p.value < 0.001 ~ "p<0.001",
      p.value < 0.01  ~ "p<0.01",
      p.value < 0.05  ~ "p<0.05",
      TRUE            ~ paste0("p=", formatC(p.value, digits = 2, format = "f"))
    )
  ) %>%
  # Put DR on top
  dplyr::mutate(isotype = factor(isotype, levels = c("HLA-DR (High vs Low)",
                                             "HLA-DP (High vs Low)",
                                             "HLA-DQ (High vs Low)"))) %>%
  dplyr::arrange(isotype)

# Choose x-limits that look nice for your effect sizes
xlims <- c(0.2, 2.5)

p <- ggplot(td, aes(x = estimate, y = isotype)) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), linewidth = 0.6) +
  scale_x_log10(limits = xlims) +
  labs(
    x = "Hazard ratio (log scale)",
    y = NULL,
    title = "HLA-II isotype groups (multivariable Cox model)",
    subtitle = "Points = HR; bars = 95% CI"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title.position = "plot",
    axis.text.y = element_text(size = 11)
  ) +
  # p-value text on the right side of each row (optional)
  geom_text(
    aes(x = xlims[2], label = p_label),
    hjust = 1, size = 3.5
  )

p

```

# DE between HLA-II groups in TCGA

## DE
```{r, eval=T}
#change IDs into gene names
hugo <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(counts_mel),
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

# 3. Subset out NAs and rename
keep <- !is.na(hugo)
counts_hugo <- counts_mel[keep, , drop=FALSE]
rownames(counts_hugo) <- hugo[keep]

rownames(counts_hugo)[which(duplicated(rownames(counts_hugo)))]

dds_all <- DESeqDataSetFromMatrix(countData = counts_hugo[, clin_final$barcode],
                              colData = subset(clin_final, barcode %in% colnames(counts_hugo)),
                              design = ~ group + Immune_infiltrate)
dds_all$group <- dds_all$group %>% relevel(ref = "Low")
keep <- rowSums(counts(dds_all)) >= 10
dds_all <- dds_all[keep,]
dds_all <- DESeq(dds_all)
plotCounts(dds_all, gene="CIITA", intgroup="group")

SKCM_all_results_filter <- as.data.frame(results(dds_all,cooksCutoff = T, name = "group_High_vs_Low")) %>% 
  dplyr::filter(baseMean >3000)
library(ggrepel)
SKCM_all_results_filter$gene <- rownames(SKCM_all_results_filter)
ggplot(SKCM_all_results_filter, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.01 & abs(log2FoldChange) > 2, "significant", "non-significant")), alpha = 0.6, size=3) +
  geom_text_repel(
    aes(label = ifelse(padj < 0.01 & abs(log2FoldChange) > 2, as.character(gene), "")),
    size = 5,
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.5, "lines")
  ) +
  theme_minimal() +
  labs(title = "SKCM dataset - MHCII+ vs MHCII-",
       x = "Log2 Fold Change",
       y = "-log10(p-value)",
       color = "Gene Significance") +
  scale_color_manual(values = c("grey", "red"))
```

## GSEA
```{r, eval=T}
gene_list_skcm <- SKCM_all_results_filter$log2FoldChange
names(gene_list_skcm) <- SKCM_all_results_filter$gene
gene_list_skcm <- sort(gene_list_skcm, decreasing = TRUE)

msigdb_hallmark <- msigdbr(species = "Homo sapiens", category = "H")
msigdb_hallmark_list <- msigdb_hallmark[, c("gs_name", "gene_symbol")]

gsea_results_skcm <- GSEA(
  geneList = gene_list_skcm,           # Ranked gene list
  TERM2GENE = msigdb_hallmark_list,          # Hallmark gene sets in the correct format
  pvalueCutoff = 0.5,            # P-value cutoff for significance
  minGSSize = 10,                 # Minimum gene set size
  maxGSSize = 500,                # Maximum gene set size
  verbose = TRUE
)

gsea_results_skcm@result$ID <- sub("HALLMARK_", replacement = "", x = gsea_results_skcm@result$ID)
gsea_results_skcm@result$regulation <- ifelse(gsea_results_skcm@result$NES > 0, "Upregulated", "Downregulated")
ggplot(subset(gsea_results_skcm@result,p.adjust<0.05), aes(x = reorder(ID, NES), y = NES)) +
  geom_bar(stat = "identity", aes(fill = regulation), width = 0.8) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  scale_fill_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
  coord_flip() +  # Flip coordinates for better readability (pathways on the y-axis)
  theme_minimal() +
  labs(title = "GSEA Waterfall Plot - HLA+ vs HLA-",
       x = "Pathway",
       y = "Normalized Enrichment Score (NES)",
       fill = "Regulation") +
  theme(axis.text.y = element_text(size = 10, face = "bold"))  

#performing progeny analysis as in the tcl data
allgenes <- unique(rownames(vst(dds_all)))
HLA_DR_genes <- grep("^HLA-DR", allgenes, value = T)
HLA_DP_genes <- grep("^HLA-DP", allgenes, value = T)
HLA_DQ_genes <- grep("^HLA-DQ", allgenes, value = T)
HLA_DO_genes <- grep("^HLA-DO", allgenes, value = T)
HLA_DM_genes <- grep("^HLA-DM", allgenes, value = T)
HLAII_genes <- unique(c("CIITA", "CD74", HLA_DR_genes,HLA_DP_genes,HLA_DQ_genes,HLA_DO_genes,HLA_DM_genes))
expr_skcm_norm <- assay(vst(dds_all))
expr_skcm_norm <- expr_skcm_norm[!(rownames(expr_skcm_norm) %in% HLAII_genes), , drop = FALSE]

activ_path <- progeny(expr_skcm_norm,
                    scale = TRUE,
                    organism = "Human",
                    top = 500,          # try 100–500 in sensitivity checks
                    perm = 300,           # no permutation; we’ll test with limma
                    return_assay = TRUE)


# 2) Differential pathway activity: HLA_high vs HLA_low (adjust for dataset)
clin_progeny = subset(clin_final, barcode %in% colnames(counts_hugo))
clin_progeny$group <- relevel(as.factor(clin_progeny$group),ref = "Low")
design <- model.matrix(~ group, data = clin_progeny)
fit <- lmFit(t(activ_path), design)            # pathways × samples -> transpose to samples × pathways
fit <- eBayes(fit)
progeny_res_skcm <- topTable(fit, coef = "groupHigh", number = Inf) %>%
  tibble::rownames_to_column("Pathway")

sig_skcm <- progeny_res_skcm %>% dplyr::filter(adj.P.Val < 0.05)
# logFC > 0 => higher activity in HLA_high
ggplot(sig_skcm, aes(x = reorder(Pathway, logFC), y = logFC, fill = logFC > 0)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "red", `FALSE` = "blue"), guide = "none") +
  labs(x = "Pathway", y = "Activity (logFC)", title = "PROGENy pathway activity: HLA+ vs HLA−")

```

# Test dedifferentiation status vs signature in TCGA-SKCM (Exploratory, not included)

```{r, eval=T}
#check differentiation status on the TCGA biopsies

expr_mat <- as.matrix(assay(vst(dds_all)))
mode(expr_mat) <- "numeric"
# load gene sets and convert to list

melanoma_dediff <- read.csv("../public/Melanoma_dediff_gene_set.csv",
                            header = TRUE, stringsAsFactors = FALSE)

# Try to standardize column names
if (!all(c("term","gene") %in% names(melanoma_dediff))) {
  if (ncol(melanoma_dediff) >= 2) {
    colnames(melanoma_dediff)[1:2] <- c("gene","term")
    melanoma_dediff <- melanoma_dediff[, c("term","gene")]
  }
}

# Make a TERM -> vector-of-genes list
melanoma_sets <- split(melanoma_dediff$gene, melanoma_dediff$term)

# Keep only genes present in expr
genes_in_expr <- rownames(expr_skcm)
melanoma_sets <- lapply(melanoma_sets, function(g) intersect(unique(g), genes_in_expr))

# Drop any empty sets (no overlap)
melanoma_sets <- melanoma_sets[lengths(melanoma_sets) > 0]

# run ssGSEA with GSVA

# (Optional) de-duplicate genes if needed (keep the most variable probe/gene)
if (any(duplicated(rownames(expr_mat)))) {
  expr_mat <- expr_mat[!duplicated(rownames(expr_mat)), ]
}

gsparam <- ssgseaParam(exprData = expr_mat, geneSets = melanoma_sets)
ssgsea_scores <- GSVA::gsva(param = gsparam,verbose = TRUE)

# ssgsea_scores: matrix [pathways x samples]

# ---- 3) heatmap of ssGSEA scores per sample/cell line ----

preferred_order <- c("Melanocytic", "Transitory", "Neural_crest-like", "Undifferentiated")
row_order <- intersect(preferred_order, rownames(ssgsea_scores))
other_rows <- setdiff(rownames(ssgsea_scores), row_order)
ssgsea_scores <- ssgsea_scores[c(row_order, sort(other_rows)), , drop = FALSE]

# Row-scale to emphasize variation across samples (nice for visualization)
ssgsea_z <- t(scale(t(ssgsea_scores)))   # z-score per gene set (row)

# If you have sample annotations (e.g., cell line, HLA status), add here:
# sample_annot <- data.frame(
#   CellLine = ...,   # a vector named by colnames(expr_mat)
#   HLAclass = ...
# )
# rownames(sample_annot) <- colnames(ssgsea_z)
annots <- clin_final$group
names(annots) <- clin_final$sample
ann_colors <- list(
  HLA_status = c(
    Low        = "#0384fc",
    High   = "red"
  )
)

levels4 <- c("Low", "High")  
annot_df <- data.frame(
  HLA_status = factor(annots[colnames(ssgsea_z)], levels = levels4)
)
rownames(annot_df) <- colnames(ssgsea_z)

pheatmap(ssgsea_z,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "none",cutree_cols = 2,
         border_color = NA,color = viridis::inferno(100),
         main = "Enrichment of melanoma differentiation stages", annotation_col = annot_df,annotation_colors =ann_colors, 
         # annotation_col = sample_annot,   # uncomment if you created sample_annot
         fontsize_row = 10,
         fontsize_col = 8, show_colnames = F)

stopifnot(identical(rownames(annot_df), colnames(ssgsea_scores)))
grp <- droplevels(annot_df$HLA_status)          # e.g., c("Low","High")

# ---- 1) group means (pathways x groups) ----
G <- model.matrix(~ 0 + grp)                    # samples x groups (0/1)
colnames(G) <- levels(grp)
G_norm <- sweep(G, 2, colSums(G), "/")          # columns sum to 1
ssgsea_group_mean <- ssgsea_scores %*% G_norm   # pathways x groups (mean per group)

# ---- 2) row z-score of the group means (what you showed) ----
ssgsea_group_z <- t(scale(t(ssgsea_group_mean)))

# undiff_biopsies <- names(which(ssgsea_scores["Melanocytic",] > ssgsea_group_z["Melanocytic",]))
# 
# clin_final$undifferentiated <- ifelse(clin_final$sample %in% undiff_biopsies, "Undifferentiated", "Non-Undifferentiated")
# 
# fit_cox <- survfit(
#   Surv(DSS.time, DSS) ~ group,
#   rbind(subset(clin_final, undifferentiated == "Non-Undifferentiated" & group == "Low"),
#                subset(clin_final, group == "High"))
# )
# 
# # 5) Plot
# ggsurvplot(
#   fit_cox,
#   data       = clin_final,
#   pval       = TRUE,
#   risk.table = TRUE
# )

#now let's classify each with one state
states <- c("Melanocytic","Transitory","Neural_crest-like","Undifferentiated")
stopifnot(all(states %in% rownames(ssgsea_scores)))

# 1) row-z the four state scores (makes them comparable)
S      <- ssgsea_scores[states, , drop = FALSE]
S_z    <- t(scale(t(S)))              # rows = states, cols = samples

# 2) seed centroids by a quick max (only to initialize centroids)
seed_call <- apply(S_z, 2, function(x) states[which.max(x)])
centroids <- sapply(states, function(st) {
  idx <- which(seed_call == st)
  if (length(idx) < 3) {               # fallback if very few seeds
    # use the theoretical unit basis as a gentle prior
    u <- rep(0, length(states)); u[match(st, states)] <- 1
    return(u)
  }
  rowMeans(S_z[, idx, drop = FALSE])
})
# centroids is 4 (dims) x 4 (states), in the given order

# 3) build a smooth ordered curve through the centroids
t_knots <- 0:3                             # 0..3 correspond to the ordered states
sf_list <- lapply(1:4, function(d) splinefun(t_knots, centroids[d, ], method = "natural"))
curve_point <- function(t) vapply(sf_list, function(f) f(t), numeric(1))  # 4D point on curve

# 4) project each sample onto the curve (find t* minimizing distance)
proj_tau <- sapply(1:ncol(S_z), function(j) {
  x <- S_z[, j]
  f <- function(t) sum((x - curve_point(t))^2)
  stats::optimize(f, interval = c(0, 3))$minimum
})
tau <- proj_tau / 3                       # 0..1 pseudotime along M→T→NC→U

# 5) assign nearest phase; flag near-boundary as Ambiguous
knots     <- c(0, 1/3, 2/3, 1)
labels    <- states
nearest_k <- sapply(tau, function(t) which.min(abs(t - knots)))
call      <- labels[nearest_k]

# boundary margin (in tau units); 0.05 ≈ “within ~5% of the trajectory”
margin <- 0.1
is_boundary <- sapply(tau, function(t) min(abs(t - knots)) < margin)
call[is_boundary] <- paste0(call[is_boundary], " (ambiguous)")

# results
tsoi_call <- factor(call, levels = c(paste0(states, " (ambiguous)"), states))
names(tsoi_call) <- colnames(S_z)
tsoi_tau  <- setNames(tau, colnames(S_z))

# quick check: counts and ordering
table(tsoi_call)

# (optional) order samples by tau and plot the 4 rows as a heatmap
ord <- order(tsoi_tau)
ann <- data.frame(Tsoi_state = sub(" \\(ambiguous\\)", "", as.character(tsoi_call)))
rownames(ann) <- names(tsoi_tau)

pheatmap::pheatmap(
  S_z[, ord, drop = FALSE],
  cluster_rows = TRUE, cluster_cols = FALSE,
  gaps_col = cumsum(table(ann$Tsoi_state[ord])),
  annotation_col = ann,
  main = "Tsoi states (row-z) ordered by trajectory (tau)",
  show_colnames = FALSE
)


clin_final$Tsoi_state <- ann$Tsoi_state[match(rownames(ann), clin_final$sample)]

#add also the continuous variable
clin_final <- clin_final %>% mutate(Melanocytic = S_z["Melanocytic",clin_final$sample ])
clin_final <- clin_final %>% mutate(Transitory = S_z["Transitory",clin_final$sample ])
clin_final <- clin_final %>% mutate(Neural_crest_like = S_z["Neural_crest-like",clin_final$sample])
clin_final <- clin_final %>% mutate(Undifferentiated = S_z["Undifferentiated",clin_final$sample ])


states <- c("Melanocytic","Transitory","Neural_crest-like","Undifferentiated")
S      <- ssgsea_scores[states, , drop = FALSE]

# row-z if you want to equalize dynamic range
S_z <- t(scale(t(S)))

# Hard call as the max-scoring state
tsoi_call <- apply(S_z, 2, function(x) states[which.max(x)])

# Optional: "Mixed" if top two are too close
# tsoi_call2 <- apply(S_z, 2, function(x) {
#   ord <- order(x, decreasing = TRUE)
#   if ((x[ord[1]] - x[ord[2]]) < 0.2) return("Mixed")
#   states[ord[1]]
# })


clin_final$Tsoi_state <- tsoi_call[match(names(tsoi_call), clin_final$sample)]
```

## Multivariate model with Tsoi states

```{r}
library(splines)
#Test with HLA-DRA, HLA-DPA1, HLA-DQA1
###HLA-DR
clin_final2$Melanocytic <- clin_final$Melanocytic[match(clin_final2$sample, clin_final$sample)]
clin_final2$Transitory <- clin_final$Transitory[match(clin_final2$sample, clin_final$sample)]
clin_final2$Neural_crest_like <- clin_final$Neural_crest_like[match(clin_final2$sample, clin_final$sample)]
clin_final2$Undifferentiated <- clin_final$Undifferentiated[match(clin_final2$sample, clin_final$sample)]

clin_final2$undifferentiated_score <- clin_final$undifferentiated_score[match(clin_final2$sample, clin_final$sample)]

clin_final2$Tsoi_state <- clin_final$Tsoi_state[match(clin_final2$sample, clin_final$sample)]

clin_final2 <- clin_final2 %>% 
  dplyr::mutate(Undifferentiated = case_when(
  Tsoi_state %in% c("Melanocytic", "Transitional", "Neural_crest-like") ~ "No",
  Tsoi_state %in% c("Undifferentiated") ~ "Yes"
))

clin_final2$Tsoi_state <- factor(clin_final2$Tsoi_state, levels = c(
  "Melanocytic", "Transitory", "Neural_crest-like","Undifferentiated")
)

# 2) Fit the Cox model with the new categorical vars
cox_mod <- coxph(
  Surv(DSS.time, DSS) ~ 
    HLA_const     + 
    Immune_infiltrate +
    group_DR +
    group_DP +
    group_DQ +
    Tsoi_state+
    Age_group   + 
    Sex      + 
    Stage_group,
  data = clin_final2
)

cox_mod_part <- coxph(
  Surv(DSS.time, DSS) ~ 
    HLA_const     + 
    Immune_infiltrate +
    Tsoi_state+
    Age_group   + 
    Sex      + 
    Stage_group,
  data = clin_final2
)
summary(cox_mod)
ggforest(cox_mod, data=clin_final2)
```


# Exploratory analyses on TIL Cohort

## Univariate analysis
```{r CONST+ score and univariate til, fig.width=15, fig.height=8}
# compute simple singscore
scores_til <- simpleScore(rank_data_til,
                      upSet   = up_list,
                      downSet = down_list)
rownames(scores_til) <- gsub(x= rownames(scores_til), pattern = "\\.", replacement = "-")
scores_til$sample <- rownames(scores_til)
scores_til <- subset(scores_til, sample %in% rownames(TIL_md_full))

clin_final_TIL <- subset(TIL_md_full, sample %in% scores_til$sample)
scores_til <- scores_til[order(scores_til$sample, clin_final_TIL$sample),]

clin_final_TIL$signature_score <- scores_til$TotalScore[
  match(clin_final_TIL$sample, scores_til$sample)
]

stopifnot(!any(is.na(clin_final_TIL$signature_score)))

clin_final_TIL <- clin_final_TIL %>% as.data.frame() %>%
  dplyr::mutate(group = if_else(signature_score >=
                           quantile(signature_score, 0.75),
                         "High","Low"))
clin_final_TIL <- clin_final_TIL %>% mutate(OS_event= case_when(
  OS_censored == 1 ~ 0,
  OS_censored == 0 ~ 1,
  TRUE ~ NA_integer_
))

clin_final_TIL <- clin_final_TIL %>% mutate(PFS_event= case_when(
  PFS_censored == 1 ~ 0,
  PFS_censored == 0 ~ 1,
  TRUE ~ NA_integer_
))

library(survival)
library(survminer)
# 4) Fit the survival model in one go, using the data frame
fit_cox_TIL <- survfit(
  Surv(OS.months, OS_event) ~ group,
  data = clin_final_TIL
)

# 5) Plot
 ggsurvplot(
  fit_cox_TIL,
  data       = clin_final_TIL,
  pval       = TRUE,
  risk.table = TRUE
)

fit <- survfit(
  Surv(PFS.months, PFS_event) ~ group,
  data = clin_final_TIL
)

# 5) Plot
ggsurvplot(
  fit,
  data       = clin_final_TIL,
  pval       = TRUE,
  risk.table = TRUE
)

```

# HLA-IICONST and response correlation in TIL (Exploratory, not included)

```{r,  fig.width=10, fig.height=5}
library(dplyr)

# 1) Define response
responders <- c("CR","PR","PR (NED)")
dat <- clin_final_TIL %>%
  dplyr::filter(!is.na(BOR)) %>%                       # drop missing outcome
  mutate(resp = BOR %in% responders,
         group = factor(group, levels = c("Low","High")))  # Low as reference

# 2) 2x2 table
tab <- table(dat$group, dat$resp)
tab
#           resp
# group    FALSE TRUE
#  High        ?    ?
#  Low         ?    ?

# 3) Fisher's exact test (exact p-value) + odds ratio & 95% CI
ft <- fisher.test(tab)
ft
# ft$estimate is the odds ratio (High vs Low); ft$conf.int is the 95% CI

# 4) Equality of proportions (Wald/chi-square, gives CI for difference)
prop.test(x = tab[, "TRUE"], n = rowSums(tab), correct = FALSE)

# 5) Unadjusted logistic regression (gives OR for High vs Low)
m0 <- glm(resp ~ group, family = binomial, data = dat)
summary(m0)
cbind(OR = exp(coef(m0)),
      confint.default(m0) |> exp())            # Wald CI; for small samples prefer exact/Firth


# 7) Simple plot: proportion responders with Wilson 95% CI
library(ggplot2)
library(Hmisc)  # for binconf()
pp <- dat %>%
  group_by(group) %>%
  summarise(n = n(),
            r = sum(resp),
            prop = r / n,
            ci = list(Hmisc::binconf(r, n, method = "wilson")[1,2:3])) %>%
  tidyr::unnest_wider(ci, names_sep = "_")

dat <- dat %>% dplyr::mutate(Obj_res=case_when(
  BOR %in% c("CR", "PR", "PR (NED)") ~ "Response",
  BOR %in% c("PD", "SD") ~ "No response"
))

dat <- dat %>%
  dplyr::filter(!is.na(Obj_res)) %>%
  mutate(
    Obj_res = factor(Obj_res, levels = c("No response", "Response"))
  )

ggplot(dat, aes(x = Obj_res, y = signature_score)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.6) +
  labs(
    x = NULL,
    y = "Scaled signature score",
    title = "Signature score in responders vs non-responders"
  ) +
  stat_compare_means(method = "wilcox.test")


pp <- dat %>%
  group_by(group) %>%
  summarise(
    n  = n(),
    r  = sum(resp),
    prop = r / n,
    ci = list(Hmisc::binconf(r, n, method = "wilson")[1,2:3])
  ) %>%
  tidyr::unnest_wider(ci, names_sep = "_")  # ci_1, ci_2

ggplot(pp, aes(x = group, y = prop)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_Lower, ymax = ci_Upper), width = 0.15) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Signature group",
    y = "Response rate",
    title = "Response rate by signature group (95% Wilson CI)"
  )


#Comparison among TCGA scores and TIL scores
combined <- bind_rows(
  clin_final_TIL %>%
    transmute(sample, signature_score,group, cohort = "TIL"),
  clin_final2 %>%
    transmute(sample, signature_score,group, cohort = "TCGA")
)
ggplot(combined, aes(x = signature_score, color = cohort)) +
  geom_density(linewidth = 2) +
  labs(x = "Signature score",
       y = "Density",
       title = "Distribution of signature score in TCGA vs trial")+
  theme_bw()


combined <- combined %>%
  group_by(cohort) %>%
  mutate(signature_z = as.numeric(scale(signature_score))) %>%
  ungroup()

ggplot(combined, aes(x = cohort, y = signature_z, colour = group)) +
   geom_violin(trim = FALSE) +
  geom_boxplot(position= position_dodge(0.9),width = 0.2, outlier.shape = NA) +
  geom_point(position = position_dodge(0.9), alpha = 0.5) +
  labs(x = NULL,
       y = "Signature score (z-score within cohort)",
       title = "Relative distribution of signature score")+
  geom_hline(yintercept = quantile(combined$signature_z, 0.75))


```

# TCR changes in HLA-high vs HLA-low (Exploratory, not included)
```{r Differences in the TCR repertoire between HLA_II_high and HLA_II_low}

file <- "../controlled_data/TIL/immunarch_TRB/"
immdata <- repLoad(file)

exp_vol <- repExplore(immdata$data, .method = "volume",.coding = T)

immdata$meta$Patient <- gsub("_", "", immdata$meta$Patient_Raw) 
immdata$meta$Patient <- gsub("2T1", "", immdata$meta$Patient) 
immdata$meta$Patient <- gsub("beforeTILs", "", immdata$meta$Patient) 
immdata$meta$Patient <- gsub("2aftIPIbefTIL", "", immdata$meta$Patient) 

immdata$meta <- immdata$meta %>% left_join(clin_final_TIL, by = "Patient")

vis(exp_vol, .by = c("group"), .meta = immdata$meta)
```

```{r}
save(clin_final_TIL, file = "../controlled_data/TIL/clin_final_TIL.R")
save(clin_final, file = "../controlled_data/TIL/clin_final.R")
```










